Spectral Experts for Estimating Mixtures of Linear Regressions

### 2013-06-13 doc/icml-presentation/presentation.tex
Fixed a couple of typos and bugs

\begin{frame}
  \frametitle{Related Work}
  \begin{itemize}
    \item<1-> Method of Moments [Pearson, 1894]
    \item<2-> Observable operator models
    \begin{itemize}
      \item Control Theory [Ljung, 1987]
      \item Observable operator models [Jaeger, 2000; Singh et al., 2004]
      \item \alert<2>{Hidden Markov models [Hsu et al, 2009]}
      \item Low-treewidth graphs [Parikh et al., 2012]
      \item Weighted finite state automata [Balle \& Mohri, 2012]
    \end{itemize}
     \item<3-> Parameter Estimation
  \begin{itemize}
    \item Mixture of Gaussians [Kalai/Moitra/Valiant, 2010]
    \item \alert{Mixture models, HMMs [Anandkumar/Hsu/Kakade, 2012]}
    \item Latent Dirichlet Allocation [Anandkumar/Hsu/Kakade, 2012]
    \item Stochastic block models [Anandkumar/Ge/Hsu/Kakade, 2012]
    \item Linear Bayesian networks [Anandkumar/Hsu/Javanmard/Kakade, 2012]
  \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Aside: Tensor Operations}
  \splitcolumn{%
    \begin{itemize}
      \item \tikzmark{tensoring} Tensoring
        $$x\tp{3}_{ijk} = x_i x_j x_k$$
      \item<2-> \tikzmark{innerp}Inner product
        \begin{align*}
          \innerp{A}{B} &= \sum_{ijk} A_{ijk} B_{ijk} \\
          \uncover<3->{%
            &= \innerp{\vvec{A}}{\vvec{B}} 
          }
        \end{align*}
    \end{itemize}
  }{%
  }
  \begin{canvas}
    \point{mark}{(1.0cm,0)};
    \point{tensoring-pos}{({pic cs:tensoring} -| mark)}
    % Tensoring
    \tikzcube{tensoring}{black,fill=white}{($(tensoring-pos) + (0,-3cm) + (0,0,0)$)}{1}{1}{1};
    \node at ($(tensoring) + (1.0cm,-0.3cm)$) {$=$};
    \tikzcube{v1}{black,fill=white}{($(tensoring) + (2.5cm,-0.3cm)$)}{1}{0.3}{0.3};
    \node (lbl1) at ($(v1) + (0.40cm,-0.1cm)$) {$\times$};
    \tikzcube{v2}{black,fill=white}{($(v1) + (1cm,0.3cm)$)}{0.3}{1}{0.3};
    \node at ($(lbl1) + (1cm,0.0cm)$) {$\times$};
    \tikzcube{v3}{black,fill=white}{($(v2) + (1cm,-0.3cm)$)}{0.3}{0.3}{1};

@@

\begin{frame}
  \frametitle{Example: Gaussian Mixture Model}
  \splitcolumn{%
    \begin{itemize}
      \item Generative process\tikzmark{gen}:
      \begin{align*}
        h &\sim \Mult([\pi_1, \pi_2, \cdots, \pi_k]) \\
        x &\sim \normal{\beta_h}{\sigma^2}.
      \end{align*}
      \item Moments:
      \begin{align*}
          \uncover<2->{%
          \E[x|h] &= \beta_h \\
          }
          \uncover<3->{%
          \E[x] &= \sum_h \pi_h \beta_h \\
          }
          \uncover<4->{%
          \tikzmark{m2}
          \E[x\tp{2}] &= \sum_h \pi_h (\beta_h \beta_h^T) + \sigma^2 \\
                      &= \sum_h \pi_h \beta_h{\tp{2}} + \sigma^2 \\
                      }
          \uncover<5->{%
          \tikzmark{m3}
          \E[x\tp{3}] &= \sum_h \pi_h \beta_h\tp{3} + \textrm{bias}.
          }
        \end{align*}
    \end{itemize}
  }{%
  }
  \begin{canvas}
    % The model
    \point{mark}{(1cm,0)};
    \point{start}{(1cm,1cm)}; %{pic cs:gen} -| mark)};
    \drawgen{($(start) + (0,1cm)$)};
    %\node[anchor=west] (diag1) at ($(start)$) {%
    %  \includegraphics[width=0.45\textwidth,height=2cm,keepaspectratio]{figures/gen.png}
    %};
    \node[anchor=west] (diag) at ($(start) + (1cm,0.0cm)$) {%
      \includegraphics[width=0.45\textwidth,height=3cm,keepaspectratio]{figures/mog.pdf}
    };

@@

  \splitcolumn{%
    \begin{itemize}
      \item \tikzmark{gen}$\E[x\tp{3}] = \sum_{h=1}^k \pi_h \beta_h\tp{3}$.
      \item<3-> If $\beta_h$ are orthogonal, the are eigenvectors!
        \begin{align*}
          \E[x\tp{3}](\beta_h,\beta_h) 
            %&= \sum_{h'=1}^k \pi_{h'} (\beta_{h}^T \beta_{h'})^2 \beta_{h'} \\
            %&= \sum_{h'=1}^k \pi_{h'} \delta_{hh'} \beta_{h'} \\
            &= \pi_{h} \beta_{h}.
        \end{align*}
      \item<4-> In general, whiten $\E[x\tp{3}]$ first.
    \end{itemize}
  }{%
  }
  \begin{canvas}
    \point{mark}{(1cm,0)};
    \point{start}{(1cm,1cm)}; %{pic cs:gen} -| mark)};

@@

  \splitcolumn{%
    \tikzmark{model}
    \begin{tikzpicture}
      \drawdisc{(0,0)}
    \end{tikzpicture}
    \begin{itemize}
      \item<2-> $h \sim \Mult([\pi_1, \pi_2, \cdots, \pi_k])$.
      \item<4-> $y = \beta_h^T x + \epsilon$.
    \end{itemize}
  }{%
  }
    \begin{canvas}
      % x, y
      \point{mark}{(0cm,0cm)}
      \point{model}{({pic cs:model} -| mark)}
      \node[anchor=west] at ($(model) + (0,-1cm)$) {%
      \includegraphics<1>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-0.pdf}
      \includegraphics<2>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-1.pdf}
      \includegraphics<3>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-2.pdf}
      \includegraphics<4>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-3.pdf}
      \includegraphics<5>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-4.pdf}
      \includegraphics<6>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-5.pdf}
      \includegraphics<7>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-6.pdf}
        };
    \end{canvas}
\end{frame}

@@

\begin{frame}[c]
  \frametitle{Finding Tensor Structure}
  \withrmargin{%
  \begin{align*}
    y &= \innerpp{\robustaltm<1>{\beta_h}{\underbrace{\beta_h}_{\textmg{random}}}}
                {x} + \epsilon \\
    \uncover<4-6>{%
    &= \robustaltm<4>{\innerp{\E[\beta_h]}{x}}
      {\mathmb{\ub{\innerp{\E[\beta_h]}{x}}_{\textrm{linear measurement}}}} 
    + \robustaltm<4-5>{\innerp{(\beta_h - \E[\beta_h])}{x} + \epsilon}
    {\mathmr{\ub{\innerp{(\beta_h - \E[\beta_h])}{x} + \epsilon}_{\textrm{noise}}}} \\
    }
  \end{align*}
  }{%
    \uncover<3->{%
        $\mboxg{mom}{\E[\beta_h] = \sum_h \pi_h \beta_h}$.
    }
  }
\end{frame}

@@

  \splitcolumn{%
  \begin{align*}
    \action<1->{%
    y\tikzmark{regA} &= \mathmb{\ob{\innerp{\E[\beta_h]}{x}}^{\textrm{linear measurement}}} &&+ \mathmr{\ob{(\beta_h - \E[\beta_h])^T x + \epsilon}^{\textrm{noise}}} \\[4ex]
    }
    \action<2->{%
    y^2 \tikzmark{regB}
      &= \left(\innerp{\beta_h}{x} + \epsilon\right)^2 && \\
    }
    \action<3->{%
    &= 
    \color{blue}
      \robustaltm<3>{%
            \innerp{\E[\beta_h\tp{2}]}{x\tp{2}}
        }{%
          \ub{\innerp{\E[\beta_h\tp{2}]}{x\tp{2}}}_{M_2} 
        }
        &&+ \color{DarkGreen} \textrm{bias} + \color{red} \textrm{noise} \\[4ex]
    }
    \action<5->{%
    y^3 \tikzmark{regC}
    &= \color{blue} \ub{\innerp{\E[\beta_h\tp{3}]}{x\tp{3}}}_{M_3} &&+ \color{DarkGreen} \textrm{bias} + \color{red} \textrm{noise} 
    }
  \end{align*}
  }{%
  }
  \begin{canvas}
    \point{mark}{(3cm,0)};
    \uncover<1-> {%
      \regressionA{($({pic cs:regA} -| mark) + (0,-2.5cm)$)};
    }
    \uncover<3-> {%
      \regressionB{($({pic cs:regB} -| mark) + (0,-3cm)$)};
    }
    \uncover<5-> {%
      \regressionC{($({pic cs:regC} -| mark) + (0,-3cm)$)};
    }
  \end{canvas}

@@

    \begin{align*}
      \robustaltm<-2>{\hat M_2}{\mathmb{\hat M_3}} &= \arg\min_{M} \sum_{(x,y)\in\mathcal{D}} \left(y^2 - 
      \robustaltm<-2>{\innerp{M}{x\tp{2}}}{\mathmb{\innerp{M}{x\tp{3}}}}
      - \textrm{bias}\right)^2 
      \uncover<2->{+
      \robustaltm<2>{\mathmb{\ub{\|M\|_{*}}_{\sum_i \sigma_i(M)}}}
        {\|M\|_{*}}
      } \\
    \end{align*}

  \begin{canvas}
    % x, y
    \alt<-2>{%
      \matrixfactorization{(-3cm,-2cm)}
    }{%
      \tensorfactorization{(-3cm,-2cm)}
    }
  \end{canvas}
\end{frame}

@@

  \begin{canvas}
    \input{spectral-experts.tikz}
    \node[style=txt] at (reg-label) {{\color{blue}\bf low-rank} regression};
    \node[style=txt] at (tf-label) {tensor factorization};

@@

  \begin{canvas}
    % x, y
    \node[anchor=west] (em) at (-6.5cm,-0.0cm) {%
    \includegraphics<1>[width=\textwidth,height=5cm,keepaspectratio]{figures/1833.pdf}
      \includegraphics<2-3>[width=\textwidth,height=5cm,keepaspectratio]{figures/EM-1833.pdf}
      \includegraphics<4-5>[width=\textwidth,height=5cm,keepaspectratio]{figures/Spectral-1833.pdf}
      \includegraphics<6->[width=\textwidth,height=5cm,keepaspectratio]{figures/Spectral-Spectral+EM-1833.pdf}
      };

    \node<1-2>[anchor=west, right=1 cm of em.east] {%
        $y = \beta^T 
            \begin{bmatrix} 
              1 \\
              t \\
              t^4 \\
              t^7
            \end{bmatrix} + \epsilon$
      };
    % x, y
    \node[anchor=west, right=6.2cm of em] at (em.west) {%
      \includegraphics<3-4>[width=\textwidth,height=5cm,keepaspectratio]{figures/EM-hist.pdf}
      \includegraphics<5>[width=\textwidth,height=5cm,keepaspectratio]{figures/EM-Spectral-hist.pdf}
      \includegraphics<6>[width=\textwidth,height=5cm,keepaspectratio]{figures/EM-Spectral-Spectral+EM-hist.pdf}
    };
  \end{canvas}
\end{frame}

>>>>

\begin{frame}
  \frametitle{Related Work}
  \begin{itemize}
    \item<1-> Method of Moments [Pearson, 1894]
    \item<2-> Observable operator models
    \begin{itemize}
      \item Control Theory [Ljung, 1987]
      \item Observable operator models [Jaeger, 2000; Littman / Sutton / Singh, 2004]
      \item \alert<2>{Hidden Markov models [Hsu/Kakade/Zhang]}
      \item Low-treewidth graphs [Parikh et al., 2012]
      \item Weighted finite state automata [Balle \& Mohri, 2012]
    \end{itemize}
     \item<3-> Parameter Estimation
  \begin{itemize}
    \item Mixture of Gaussians [Kalai/Moitra/Valiant, 2010]
    \item \alert{Mixture models, HMMs [Anandkumar/Hsu/Kakade, 2012]}
    \item Latent Dirichlet Allocation [Anandkumar/Hsu/Kakade, 2012]
    \item Stochastic block models [Anandkumar/Ge/Hsu/Kakade, 2012]
    \item Linear Bayesian networks [Anandkumar/Hsu/Javanmard/Kakade, 2012]
  \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Aside: Tensor Operations}
  \splitcolumn{%
    \begin{itemize}
      \item \tikzmark{tensoring} Tensoring
        \begin{align*}
          x\tp{3}       &= x \otimes x \otimes x \\
          x\tp{3}_{ijk} &= x_i x_j x_k
        \end{align*}
      \item<2-> \tikzmark{innerp}Inner product
        \begin{align*}
          \innerp{A}{B} &= \sum_{ijk} A_{ijk} B_{ijk} \\
          \uncover<3->{%
            &= \innerp{\vvec{A}}{\vvec{B}} 
          }
        \end{align*}
    \end{itemize}
  }{%
  }
  \begin{canvas}
    \point{mark}{(1.0cm,0)};
    \point{tensoring-pos}{({pic cs:tensoring} -| mark)}
    % Tensoring
    \tikzcube{tensoring}{black,fill=white}{($(tensoring-pos) + (0,-3cm) + (0,0,0)$)}{1}{1}{1};
    \node at ($(tensoring) + (1.0cm,-0.3cm)$) {$=$};
    \tikzcube{v1}{black,fill=white}{($(tensoring) + (2.5cm,-0.3cm)$)}{1}{0.3}{0.3};
    \node (lbl1) at ($(v1) + (0.40cm,-0.1cm)$) {$\times$};
    \tikzcube{v2}{black,fill=white}{($(v1) + (1cm,0.3cm)$)}{0.3}{1}{0.3};
    \node at ($(lbl1) + (1cm,0.0cm)$) {$\times$};
    \tikzcube{v3}{black,fill=white}{($(v2) + (1cm,-0.3cm)$)}{0.3}{0.3}{1};

@@

\begin{frame}
  \frametitle{Example: Gaussian Mixture Model}
  \splitcolumn{%
    \begin{itemize}
      \item Generative process\tikzmark{gen}:
      \begin{align*}
        h &\sim \Mult([\pi_1, \pi_2, \cdots, \pi_k]) \\
        x &\sim \normal{\beta_h}{\sigma^2}.
      \end{align*}
      \item<2-> Moments:
      \begin{align*}
          \uncover<2->{%
          \E[x|h] &= \beta_h \\
          }
          \uncover<3->{%
          \E[x] &= \sum_h \pi_h \beta_h \\
          }
          \uncover<4->{%
          \tikzmark{m2}
          \E[x\tp{2}] &= \sum_h \pi_h (\beta_h \beta_h^T) + \sigma^2 \\
                      &= \sum_h \pi_h \beta_h{\tp{2}} + \sigma^2 \\
                      }
          \uncover<5->{%
          \tikzmark{m3}
          \E[x\tp{3}] &= \sum_h \pi_h \beta_h\tp{3} + \textrm{bias}.
          }
        \end{align*}
    \end{itemize}
  }{%
  }
  \begin{canvas}
    % The model
    \point{mark}{(1cm,0)};
    \point{start}{(1cm,1cm)}; %{pic cs:gen} -| mark)};
    \drawgen{($(start) + (0,1cm)$)};
    %\node[anchor=west] (diag1) at ($(start)$) {%
    %  \includegraphics[width=0.45\textwidth,height=2cm,keepaspectratio]{figures/gen.png}
    %};
    \node[anchor=west] (diag) at ($(start) + (1cm,0.0cm)$) {%
      \includegraphics[width=0.45\textwidth,height=3cm,keepaspectratio]{figures/mog.pdf}
    };

@@

  \splitcolumn{%
    \begin{itemize}
      \item \tikzmark{gen}$\E[x\tp{3}] = \sum_{h=1}^k \pi_h \beta_h\tp{3}$.
      \item<3-> If $\beta_h$ are orthogonal, they are eigenvectors!
        \begin{align*}
          \E[x\tp{3}](\beta_h,\beta_h) 
            %&= \sum_{h'=1}^k \pi_{h'} (\beta_{h}^T \beta_{h'})^2 \beta_{h'} \\
            %&= \sum_{h'=1}^k \pi_{h'} \delta_{hh'} \beta_{h'} \\
            &= \pi_{h} \beta_{h}.
        \end{align*}
      \item<4-> In general, whiten $\E[x\tp{3}]$ first.
    \end{itemize}
  }{%
  }
  \begin{canvas}
    \point{mark}{(1cm,0)};
    \point{start}{(1cm,1cm)}; %{pic cs:gen} -| mark)};

@@

  \splitcolumn{%
    \tikzmark{model}
    \begin{tikzpicture}
      \drawdisc{(0,0)}
    \end{tikzpicture}
    \begin{itemize}
      \item<2-> Given x
      \begin{itemize}
        \item<2-> $h \sim \Mult([\pi_1, \pi_2, \cdots, \pi_k])$.
        \item<3-> $y = \beta_h^T x + \epsilon$.
      \end{itemize}
    \end{itemize}
  }{%
  }
    \begin{canvas}
      % x, y
      \point{mark}{(0cm,0cm)}
      \point{model}{({pic cs:model} -| mark)}
      \node[anchor=west] at ($(model) + (0,-1cm)$) {%
      \includegraphics<1>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-0.pdf}
      \includegraphics<2>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-1.pdf}
      \includegraphics<3>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-2.pdf}
      \includegraphics<4>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-3.pdf}
      \includegraphics<5>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-4.pdf}
      \includegraphics<6>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-5.pdf}
      \includegraphics<7>[width=5cm,height=6cm,keepaspectratio]{figures/mlr-6.pdf}
        };
    \end{canvas}
\end{frame}

@@

\begin{frame}[c]
  \frametitle{Finding Tensor Structure}
  \withrmargin{%
  \begin{align*}
    y &= \innerpp{\robustaltm<1>{\beta_h}{\underbrace{\beta_h}_{\textmg{random}}}}
                {x} + \epsilon \\
    \uncover<3-5>{%
    &= \robustaltm<3>{\innerp{\E[\beta_h]}{x}}
      {\mathmb{\ub{\innerp{\E[\beta_h]}{x}}_{\textrm{linear measurement}}}} 
    + \robustaltm<3-4>{\innerp{(\beta_h - \E[\beta_h])}{x} + \epsilon}
    {\mathmr{\ub{\innerp{(\beta_h - \E[\beta_h])}{x} + \epsilon}_{\textrm{noise}}}} \\
    }
  \end{align*}
  }{%
    \uncover<3->{%
        $\mboxg{mom}{\E[\beta_h] = \sum_h \pi_h \beta_h}$.
    }
  }
\end{frame}

@@

  \splitcolumn{%
  \begin{align*}
    \action<1->{%
    y\tikzmark{regA} &= \mathmb{\ob{\innerp{\E[\beta_h]}{x}}^{\textrm{linear measurement}}} &&+ \mathmr{\ob{(\beta_h - \E[\beta_h])^T x + \epsilon}^{\textrm{noise}}} \\[4ex]
    }
    \action<2->{%
    y^2 \tikzmark{regB}
      &= \left(\innerp{\beta_h}{x} + \epsilon\right)^2 && \\
    }
    \action<3->{%
    &= 
    \color{blue}
      \robustaltm<3>{%
            \innerpp{\E[\beta_h\tp{2}]}{x\tp{2}}
        }{%
          \innerpp{\ub{\E[\beta_h\tp{2}]}_{M_2}}{x\tp{2}}
        }
        &&+ \color{DarkGreen} \textrm{bias} + \color{red} \textrm{noise} \\[4ex]
    }
    \action<5->{%
    y^3 \tikzmark{regC}
    &= \color{blue} \innerpp{\ub{\E[\beta_h\tp{3}]}_{M_3}}{x\tp{3}} &&+ \color{DarkGreen} \textrm{bias} + \color{red} \textrm{noise} 
    }
  \end{align*}
  }{%
  }
  \begin{canvas}
    \point{mark}{(3cm,0)};
    \uncover<1-> {%
      \regressionA{($({pic cs:regA} -| mark) + (0,-2.5cm)$)};
    }
    \uncover<3-> {%
      \regressionB{($({pic cs:regB} -| mark) + (0,-3cm)$)};
    }
    \uncover<5-> {%
      \regressionC{($({pic cs:regC} -| mark) + (0,-3cm)$)};
    }
  \end{canvas}

@@

    \begin{align*}
      \robustaltm<-3>{\hat M_2}{\mathmb{\hat M_3}} &= \arg\min_{M} \sum_{(x,y)\in\mathcal{D}} \left(y^2 - 
      \robustaltm<-3>{\innerp{M}{x\tp{2}}}{\mathmb{\innerp{M}{x\tp{3}}}}
      - \textrm{bias}\right)^2 
      \uncover<3->{+
      \robustaltm<2>{\mathmb{\ub{\|M\|_{*}}_{\sum_i \sigma_i(M)}}}
        {\|M\|_{*}}
      } \\
    \end{align*}

  \begin{canvas}
    % x, y
    \uncover<2-3>{
      \matrixfactorization{(-3cm,-2cm)}
    }
    \only<4>{
      \tensorfactorization{(-3cm,-2cm)}
    }
  \end{canvas}
\end{frame}

@@

  \begin{canvas}
    \input{spectral-experts.tikz}
    \node[style=txt] at (reg-label) {\robustalt<1>{\color{blue}\bf low-rank}{low-rank} regression};
    \node[style=txt] at (tf-label) {tensor factorization};

@@

  \begin{canvas}
    % x, y
    \node[anchor=west] (em) at (-6.5cm,-0.0cm) {%
    \includegraphics<1>[width=\textwidth,height=5cm,keepaspectratio]{figures/1833.pdf}
      \includegraphics<2-3>[width=\textwidth,height=5cm,keepaspectratio]{figures/EM-1833.pdf}
      \includegraphics<4>[width=\textwidth,height=5cm,keepaspectratio]{figures/Spectral-1833.pdf}
      \includegraphics<5->[width=\textwidth,height=5cm,keepaspectratio]{figures/Spectral-Spectral+EM-1833.pdf}
      };

    \node<1-2>[anchor=west, right=1 cm of em.east] {%
        $y = \beta^T 
            \ub{\begin{bmatrix} 
              1 \\
              t \\
              t^4 \\
              t^7
            \end{bmatrix}}_{x} + \epsilon$
      };
    % x, y
    \node[anchor=west, right=6.2cm of em] at (em.west) {%
      \includegraphics<3>[width=\textwidth,height=5cm,keepaspectratio]{figures/EM-hist.pdf}
      \includegraphics<4>[width=\textwidth,height=5cm,keepaspectratio]{figures/EM-Spectral-hist.pdf}
      \includegraphics<5>[width=\textwidth,height=5cm,keepaspectratio]{figures/EM-Spectral-Spectral+EM-hist.pdf}
    };
  \end{canvas}
\end{frame}
